---
title: "Personal Finance"
output: github_document
---

Since 2011 I have been keeping meticulous records of everything I spend money on. Even for example the 6p Carrot I bought at 19:47 on the 2nd of August 2014 at Tesco (store id 2990)! There's a total of `r nrow(unique(Finance[,c("Date","Time")]))` recipts covering a duration of `r round(as.numeric((max(Finance$Date, na.rm=T) - min(Finance$Date, na.rm=T))/365),1)` years.

```{r load_librarys, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, fig.width=10, fig.asp=1/1.5, fig.retina=2, message=FALSE, warning=FALSE)
library(data.table, quietly = TRUE)
library(ggplot2)
library(lubridate)
library(dplyr)
library(scales)
library(ggTimeSeries) 
#library(plotly)
```

Integer values (£1,£2 etc.) are more common, as well as an increase in the frequency of £X.50 and £X.25 etc values.

```{r}
undergrad <- fread("Finance - Undergrad Expenses.tsv")
dphil <- fread("Finance - DPhil Expenses.tsv")

Finance <- rbind(undergrad[,-c("Day","Notes"), with=FALSE], dphil[,-c("Totals:","V13"), with=FALSE])
Finance <- Finance[Item!=""] # remove empty lines

# Reformat Date and Time
Finance$Date <- dmy(Finance$Date)
#Finance$Time <- hms(substr(Finance$Time,0,5))
Finance$Time <- as.POSIXct(substr(Finance$Time,0,5),tz="GMT",format='%H:%M') #, origin="1970-01-01"
Finance[,Term:= factor(Term, levels=c("Michaelmas","Christmas","Hilary","Trinity","Summer"))]
Finance[,Category:= as.factor(Category)]
Finance[, Percentage := Price / sum(Price, na.rm = TRUE)]

# Distribution of Item Prices
ggplot(Finance, aes(Price)) +
  geom_histogram(binwidth=0.02) +
  scale_x_continuous(labels = dollar_format(prefix = "£"), limits=c(0,5)) +
  labs(x="Price", y="Number of Items")
```

Looking at the daily spend over time clearly shows the periods I was at university and the periods I was at home (with food etc. paid for by my parents) by the gaps where daily spending is £0.

The vast majority of days I spent less than £10 (excluding accommodation). Note that on days when I spent nothing I did not record anything so these do not appear in the graph as a spike at £0.

```{r fig.width=6}
# Aggregate: Price Per Day
FinanceByDay <- Finance[,.(Spend = sum(Price, na.rm=T), Term=unique(Term)), by = Date]

ggplot_calendar_heatmap(FinanceByDay[!is.na(Date)], 'Date', 'Spend') +
  theme(legend.position = "bottom", legend.key.width=unit(3,"cm")) +
  scale_fill_distiller(palette = 'YlOrRd', direction = 1, trans = "log10", breaks=c(1,5,10,50,100,500)) +
  facet_wrap(~Year, ncol = 2)


# Distribution of Spend per Day
ggplot(FinanceByDay, aes(Spend)) +
  geom_histogram(binwidth=0.1) +
  xlim(0,25) + labs(x="Daily Spend (£)", y="# of Days with a given spend", title="Distribution of Daily Spend")

# Aggregate: Price Per Term
FinanceByTerm <- Finance[,.(Spend = sum(Price, na.rm = T)), by = c('Term','Year')]

# Histogram - Spend Per Term
ggplot(FinanceByTerm[Term!="Christmas"], aes(Term,Spend)) +
  geom_col() +
  theme(axis.text.x=element_text(angle=40,hjust=1,vjust=1,size=15)) +
  scale_y_continuous(labels = dollar_format(prefix = "£"))+
  labs(x="Term", y="Termly Spend") + facet_grid(~Year)
```

Breaking down my spending by category, Rent accounts for the vast majority of my spending followed by Food.

```{r}
# Aggregate: Spend Per Categrory
FinanceByCategory <- Finance[,.(Spend = sum(Percentage, na.rm = T)), by = Category]
FinanceByCategory[, Category := factor(Category, levels=FinanceByCategory[order(Spend)]$Category)]

# Histogram Spend Per Vendor
ggplot(FinanceByCategory, aes(Category, Spend)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent)+
	labs(x="Category", y="Total Spend Per Category")

```

Looking at total spend per vendor the top three vendors simply reflect my rental expenditure, followed by Tesco's and Sainsbury's where I do my grocery shopping.

```{r}
# Aggregate: Spend Per Vendor
FinanceByVendor <- Finance[,.(Spend = sum(Percentage, na.rm = TRUE)), by = Vendor]
FinanceByVendor[,Vendor := factor(Vendor, levels=FinanceByVendor[order(Spend)]$Vendor)]

# Histogram Spend Per Vendor
ggplot(FinanceByVendor[order(-Spend)][1:15], aes(Vendor,Spend)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
	labs(x="", y="Total Spend Per Vendor")

#scale_y_continuous(labels = dollar_format(prefix = "£"))+

```

My most commonly bought item is milk, bananas and bread. 

```{r}
# Aggregate: Number of items
FinanceByItem <- Finance[,.(Spend = sum(Price, na.rm=T),.N), by = Item][order(-N)]
FinanceByItem[,Item := factor(Item, levels=FinanceByItem[order(N)]$Item)]

# Histogram Number of items
ggplot(FinanceByItem[order(-N)][1:20], aes(Item,N)) +
  geom_col() +
  coord_flip() +
	labs(x="", y="Total Number of Purchases")

# to fix, banana(s), carrot(s), Potatoe(s), caps custard creams, Apple(s)
```